---
title: "Testing_2026"
author: "Kevin Ly"
date: "2026-03-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter
```{r}
library(tidyverse)
library(elo)
library(tibble)
library(FNN)
library(rpart) #For classification and regression trees
library(rattle) #For creating nice visualizations of trees

Logisticmodel <- readRDS("march_madness_model.rds")
LogisticmodelSOS <- readRDS("march_madness_model_sos.rds")
gamedata <- readRDS("gamedata_2026.rds")
careerElos <- readRDS("career_elos_2026.rds")
teams <- readRDS("teams_2026.rds")
sos <- readRDS("sos_2026.rds")

MNCAATourneySeeds <- read.csv("~/Documents/MarchMadness/2026/MNCAATourneySeeds.csv")
```

### Part 5
Start with an example prediction of the very last row from gamedata.  Do it first without SOS.
```{r}
predict(Logisticmodel, newdata = data.frame(lower_team_elo = 1306, 
                                            higher_team_elo = 1618, 
                                            lower_team_rebdiff = 17,
                                            lower_team_3diff = 4,
                                            lower_team_ftdiff = -5), type = "response")
```
Then do it with SOS

```{r}
predict(LogisticmodelSOS, newdata = data.frame(lower_team_elo = 1306, 
                                            higher_team_elo = 1618, 
                                            lower_team_rebdiff = 17,
                                            lower_team_3diff = 4,
                                            lower_team_ftdiff = -5,
                                            lower_team_sosdiff = 37), type = "response")
```


This is the probability the lower team wins so we'll have to check the team ID and assign probability based on if this team has a smaller ID than the other.  

Now we want to read in and build all of the predictions of every team vs every other team in the tourney.  This will allow us to predict future round matchups. First we need to get the final elos into a place we can use them
```{r}
gamedata2025 <- gamedata %>% filter(Season == 2026)

eloScores2025 <- elo.run(score(gamedata2025$WScore, gamedata2025$LScore) ~ gamedata2025$WTeamName + gamedata2025$LTeamName, data = gamedata2025, k = 28)

finalElos <- as.data.frame(final.elos(eloScores2025))
finalElos <- rownames_to_column(finalElos, "TeamName")
names(finalElos)[names(finalElos) == 'final.elos(eloScores2025)'] <- 'elo'
teamstats <- merge(teams[,c(1,2)], finalElos, by = "TeamName")
```

Next for every team we want to calculate the average rebounds and average 3s made per game
```{r}
# I know I'm just pulling average rebounds for just the wins as a quick metric
AVGREB <- gamedata2025 %>% group_by(WTeamID) %>% summarise_at(vars(WOR, WDR), list(avgreb = mean))
names(AVGREB)[names(AVGREB) == 'WTeamID'] <- 'TeamID'
teamstats <- merge(teamstats, AVGREB, by = "TeamID")
teamstats <- teamstats %>% mutate(totavgreb = teamstats$WOR_avgreb+ teamstats$WDR_avgreb)


# I know I'm just pulling avg 3's made for just the wins as a quick metric
AVG3 <- gamedata2025 %>% group_by(WTeamID) %>% summarise_at(vars(WFGM3), list(avg3m = mean))
teamstats <- merge(teamstats, AVG3, by.x = "TeamID", by.y = "WTeamID")

# I know I'm just pulling avg 3's made for just the wins as a quick metric
AVGFT <- gamedata2025 %>% group_by(WTeamID) %>% summarise_at(vars(WFTM), list(avgftm = mean))
teamstats <- merge(teamstats, AVGFT, by.x = "TeamID", by.y = "WTeamID")

teamstats <- merge(teamstats, sos, by = "TeamName")
```

Finally show the rankings for the teams
```{r}
teamstats %>% arrange(desc(elo))
```

Now we should test Saint Mary's against Duke and Miami OH against Virgina just to see how SOS affects outcomes

First, Miami of Ohio against Virginia

```{r}
predict(Logisticmodel, newdata = data.frame(lower_team_elo = 1727, 
                                            higher_team_elo = 1716, 
                                            lower_team_rebdiff = -5,
                                            lower_team_3diff = -1,
                                            lower_team_ftdiff = -1), type = "response")
```

A 54pct probability of beating Virginia.  Now if we add in SOS let's see how they do
```{r}
predict(LogisticmodelSOS, newdata = data.frame(lower_team_elo = 1727, 
                                            higher_team_elo = 1716, 
                                            lower_team_rebdiff = -5,
                                            lower_team_3diff = -1,
                                            lower_team_ftdiff = -1,
                                            lower_team_sosdiff = -96), type = "response")
```

A 51pct chance with current stats.  So this moves the percent chance by about 3pct when the strength of schedule is significant which is good.  