---
title: "March Madness 2026"
author: "Kevin Ly"
date: "2026-03-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter
```{r}
library(tidyverse)
library(elo)
library(tibble)
library(FNN)
library(rpart) #For classification and regression trees
library(rattle) #For creating nice visualizations of trees

teams <- read.csv("~/Documents/MarchMadness/2026/MTeams.csv")
box_scores <- read.csv("~/Documents/MarchMadness/2026/MRegularSeasonDetailedResults.csv")
conf_tourney_scores <- read.csv("~/Documents/MarchMadness/2026/MConferenceTourneyGames.csv")
tourney_scores <- read.csv("~/Documents/MarchMadness/2026/MNCAATourneyDetailedResults.csv")
MNCAATourneySeeds <- read.csv("~/Documents/MarchMadness/2026/MNCAATourneySeeds.csv")
```

The basic training strategy is going to be 
1. Data preparation, binding teams, box scores, and tourney scores into one data frame
2. Include the winning and losing team's ELO scores in our main data set.  Also use Elo to estimate the teams strength of schedule to see how that factors into wins
3. Build an average of the team's recent performance using the box scores
4. Run a logistic regression with the data from 2 and 3

## Data preparation

### Part 1
First we should append the tourney and regular season box scores together
```{r}
gamedata <- rbind(box_scores,tourney_scores)
```

Next I want to add columns that give us winning and losing team names.  Do the winning teams first
```{r}
gamedata <- merge(teams[,c(1,2)], gamedata, by.x = "TeamID", by.y = "WTeamID", all.y = TRUE)
gamedata <- gamedata %>% rename(WTeamID = TeamID,
                WTeamName = TeamName)
```

Next do the losing teams
```{r}
gamedata <- merge(teams[,c(1,2)], gamedata, by.x = "TeamID", by.y = "LTeamID", all.y = TRUE)
gamedata <- gamedata %>% rename(LTeamID = TeamID,
                LTeamName = TeamName)
```


We need to turn the game wins from a winning team ID/ losing team ID to a response that we can actually train the regression on
```{r}
gamedata <- gamedata %>% mutate(lower_team = pmin(WTeamID, LTeamID), # lower refers to ID number
           higher_team = pmax(WTeamID, LTeamID),
           lower_team_wins = ifelse(lower_team == WTeamID, "YES", "NO"))
```

Let's order by season and day number
```{r}
gamedata <- gamedata[with(gamedata, order(Season, DayNum)),]
```

### Part 2
Next we need to try to calculate the winning and losing teams ELO scores and add them to the game data.  We're going to use the elo.run function from the elos package to calculate this

The regression here is really important, it resets elo scores after each season halfway back to the 1500 starting score.  The k is also fairly important as it takes into account the recency of wins.  Higher k means the model accounts for recent wins more heavily.  Some people choose from 20-30, a lot of people will use 40 to heavily favor recency.  

```{r}
# Parameter selection based on holdout validation across 3 tournament years
# k=20 was optimal in all three years - consistent signal
# regress varied by tournament type:
#   2023 (upset heavy): regress = 0.3 optimal
#   2024 (mixed):       regress = 0.7 optimal  
#   2025 (chalk):       regress = 1.0 optimal
# Average optimal regress ~ 0.67, settled on 0.6 as principled middle ground
# All years showed log loss > 0.5 reflecting inherent tournament unpredictability
# Note: SOS feature showed marginal improvement on 2025, less so on 2024
eloScores <- elo.run(score(gamedata$WScore, gamedata$LScore) ~ gamedata$WTeamName + gamedata$LTeamName 
                     + regress(gamedata$Season,1500,0.6)
                     , data = gamedata, k = 20)

careerElos <- as.data.frame(final.elos(eloScores))
careerElos <- rownames_to_column(careerElos, "TeamName")
names(careerElos)[2] <- "elo_career"
```

Now we're going to add the elos into the gamedata so we can use it in our regression training
```{r}
gamedata <- gamedata %>% mutate(WTeamElo = eloScores$elos[,7], LTeamElo = eloScores$elos[,8])
```

We also want to flag the lower_teams_elo so we can train on it.

```{r}
gamedata <- gamedata %>% mutate(lower_team_elo = ifelse(lower_team_wins == "YES", WTeamElo, LTeamElo))
gamedata <- gamedata %>% mutate(higher_team_elo = ifelse(lower_team_wins == "YES", LTeamElo, WTeamElo))
```


Let's also add a new strength of schedule feature to the model.This feature is going to take into account the average elo of the teams opponents, this will give us some measure of if someone is from a conference with really low ranked teams.
```{r}
# Strength of schedule: average Elo of all opponents faced in 2026 season
elo_lookup <- setNames(careerElos$elo, careerElos$TeamName)

sos_wins <- gamedata %>%
  group_by(WTeamName) %>%
  summarise(sos = mean(elo_lookup[LTeamName], na.rm = TRUE), .groups = "drop") %>%
  rename(TeamName = WTeamName)

sos_losses <- gamedata %>%
  group_by(LTeamName) %>%
  summarise(sos = mean(elo_lookup[WTeamName], na.rm = TRUE), .groups = "drop") %>%
  rename(TeamName = LTeamName)

sos <- bind_rows(sos_wins, sos_losses) %>%
  group_by(TeamName) %>%
  summarise(sos = mean(sos, na.rm = TRUE), .groups = "drop")

# Then join back to gamedata on WTeamName and LTeamName separately
gamedata <- gamedata %>%
  left_join(sos, by = c("WTeamName" = "TeamName")) %>%
  rename(WTeamSOS = sos) %>%
  left_join(sos, by = c("LTeamName" = "TeamName")) %>%
  rename(LTeamSOS = sos)

# Then flag lower/higher just like you do with Elo
gamedata <- gamedata %>%
  mutate(lower_team_sos = ifelse(lower_team_wins == "YES", WTeamSOS, LTeamSOS),
         higher_team_sos = ifelse(lower_team_wins == "YES", LTeamSOS, WTeamSOS))
```


### Part 3
Here we're going to just add a few extra metrics that we know matter. We'll calculate based on fg pct, 3pt percentage, rebounding differential, and turnover differential.  
```{r}
gamedata <- gamedata %>% mutate(WFGPCT = WFGM/WFGA)
gamedata <- gamedata %>% mutate(W3PTPCT = WFGM3/WFGA3)
gamedata <- gamedata %>% mutate(WFTPCT = WFTM/WFTA )
gamedata <- gamedata %>% mutate(WREBDIFF = (WOR+WDR-LOR-LDR))
gamedata <- gamedata %>% mutate(WTODIFF = (WTO-LTO))
gamedata <- gamedata %>% mutate(W3DIFF = (WFGM3-LFGM3))
gamedata <- gamedata %>% mutate(WFTDIFF = (WFTM-LFTM))
gamedata <- gamedata %>% mutate(WteamSOSDiff = (WTeamSOS-LTeamSOS))

#Again flag the lower teams rebound differential
gamedata <- gamedata %>% mutate(lower_team_rebdiff = ifelse(lower_team_wins == "Yes", WREBDIFF, LDR+LOR-WOR-WDR))
gamedata <- gamedata %>% mutate(lower_team_todiff = ifelse(lower_team_wins == "Yes", WTODIFF, LTO-WTO))
gamedata <- gamedata %>% mutate(lower_team_3diff = ifelse(lower_team_wins == "Yes", W3DIFF, LFGM3-WFGM3))
gamedata <- gamedata %>% mutate(lower_team_ftdiff = ifelse(lower_team_wins == "Yes", WFTDIFF, LFTM-WFTM))
gamedata <- gamedata %>% mutate(lower_team_sosdiff = ifelse(lower_team_wins == "Yes", WTeamSOSDiff, LTeamSOS-WTeamSOS))
```

### Part 4
Let's train our regression
```{r}
levels(as.factor(gamedata$lower_team_wins))
gamedata <- gamedata %>% mutate(lower_team_wins_num = ifelse(lower_team_wins == "YES", 1, 0))

# I tried adding all of the WFGPCT, WFTPCT, W3PTPCT and so on and none of them were significant
#.  Only rebounding and the 3 point differential was significant because rebounding is KING!
Logisticmodel <- glm(lower_team_wins_num ~ lower_team_elo + higher_team_elo
                      + lower_team_rebdiff 
                     + lower_team_3diff 
                     + lower_team_ftdiff
                       , family = binomial, data = gamedata)
summary(Logisticmodel)
```

Let's also train the model with SOS to compare
```{r}
# I tried adding all of the WFGPCT, WFTPCT, W3PTPCT and so on and none of them were significant
#.  Only rebounding and the 3 point differential was significant because rebounding is KING!
LogisticmodelSOS <- glm(lower_team_wins_num ~ lower_team_elo + higher_team_elo
                      + lower_team_rebdiff 
                     + lower_team_3diff 
                     + lower_team_ftdiff
                     + lower_team_sosdiff
                       , family = binomial, data = gamedata)
summary(LogisticmodelSOS)
```


```{r}
saveRDS(Logisticmodel, "march_madness_model.rds")
saveRDS(LogisticmodelSOS, "march_madness_model_sos.rds")
saveRDS(gamedata, "gamedata_2026.rds")        # needed for sos and gamedata2025
saveRDS(careerElos, "career_elos_2026.rds")   # needed for elo_lookup in SOS calc
saveRDS(teams, "teams_2026.rds") 
saveRDS(sos, "sos_2026.rds")
```

